Efficient pedigree recording for fast population genetics simulation

### 2017-12-27 forwards_paper.tex
get rid of sentence starting with This
get rid of another sentence starting with This

%% Progressive memory reduction as a function of T
To get an idea of how required space depends on the length of time between simplification steps,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
This holds up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

@@

Our method stores genealogies, and so records substantially more information
than would a method only recording genotypes.
However, since the simplification algorithm requires computational effort,
it is informative to compare computational complexity of the algorithm
to one that propagates neutral genotypes.
A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
On the other hand,
in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
\plr{XXX}
The simplification algorithm is linear in the number of edges of the input tree sequence
(because it must process each edge),
and so multiplies this by a constant factor.
This implies that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
while our implementation is only $O(T N \log(2N))$.

>>>>

%% Progressive memory reduction as a function of T
To get an idea of how required space depends on the length of time between simplification steps,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

@@

Our method stores genealogies, and so records substantially more information
than would a method only recording genotypes.
However, since the simplification algorithm requires computational effort,
it is informative to compare computational complexity of the algorithm
to one that propagates neutral genotypes.
A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
On the other hand,
in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
\plr{XXX}
The simplification algorithm is linear in the number of edges of the input tree sequence
(because it must process each edge),
and so multiplies this by a constant factor.
These considerations imply that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
while our implementation is only $O(T N \log(2N))$.
